---
title: "Analysis of Spot Speed Data"
author: "ECIV 542"
date: today
format:
  html:
    toc: true            # Generates a clickable table of contents
    code-fold: true      # Hides code chunks behind a clickable toggle
    embed-resources: true # Compiles everything into a single, shareable HTML file
execute:
  echo: false
  warning: false
  error: false
---

```{r}
#| label: load library

library(readr)
library(dplyr)
library(ggplot2)
library(hrbrthemes)
library(DT)

```

# Descriptive Statistics

* Central Tendency
  - **Mean**: sum of observed values (speed) divided by number of obs
  - **Median**: value (speed) that divides the data into equal parts
  - **Mode**: value at which most number of observations occur
  - **Pace**: a 10 mph block of max observation
* Dispersion
  - standard deviation

# Problem: 
Figure 11.1 represents a typical set of field data from a spot speed study taken at a location of interest on a major arterial.

# Speed Data 

```{r}
#| label: read data

# Read the local data file
# Assumes 'traffic_data.csv' is in the same folder as this .qmd file
spdData <- read_csv("images/chap11/ch11spdData.csv")
```

# Frequency Distribution Table



```{r}
totVeh <- sum(spdData$numVeh)

spdDataStat <- spdData |> 
  mutate(freq = round(numVeh *100 / totVeh, 2),
         cumulFreq = cumsum(freq),
         nS = numVeh * spdMiddle,
         nS2 = numVeh * spdMiddle^2) 

sum_ns <- sum(spdDataStat$nS)

spdDataStat |>
  janitor::adorn_totals("row") |> 
  datatable(
  options = list(
    searching = FALSE, # Disables global searching/filtering
    pageLength = 25,   # Number of rows per page
    dom = 'tp'         # Only displays the (t)able and (p)agination controls
  ))
 
  
```
# Summary Statistics: Central Tendency

## Mean
* Sum number of observation in groups, $\sum{n_iS_i}$
* The total number of vehicle, $N$ *
* mean, $\bar{x} = \frac{\sum{n_iS_i}}{N}$
  - $N = `r totVeh`$
  - $\sum{n_iS_i} = `r sum_ns`$
  - $\bar{x} = \frac{`r sum_ns`}{`r totVeh`} = 48.2$ mph

## Meadian

* speed that divides the distribution into equal parts
  - there are as many observations of speeds higher than the median as there are lower than the median
* $N/2=368/2=184$ th observation divides the data into equal parts
* or, the speed at which $50%$ cumulative frequency is found
* From the table, $P_{50} = 47$ mph

## Mode
* Speed at which max observation occurs
* From Table: max observation is $65$ at speed $47$ mph
* Mode: $47$ mph

## Low Speed and High Speed
* Speed at which 15% drivers are driving 
* From Table: 15% occurs between speed $41$ and $43$ mph; mean $42$ mph
* From the table, $P_{15} = 42$ mph

* Speed at which 85% drivers are driving 
* From Table: 85% occurs between speed $53$ and $55$ mph; mean $54$ mph
* From the table, $P_{85} = 54$ mph

## Graphical Approach: Median, Mode, Pace

```{r}
#| label: hist graph


p1 <- spdDataStat |> 
  ggplot() +  theme_ipsum() +
  geom_rect(aes(xmin = spdLmtLower + 0.15, xmax = spdLmtUpper, ymin = 0, ymax = numVeh), fill = "black")+
  geom_text(aes(x=spdMiddle, y= numVeh, label = numVeh), vjust = 0.50, hjust = -0.50, size = 4, angle = 90) +
  scale_x_continuous(limits = c(30, 70), breaks = seq(30, 70, by = 2)) +
  scale_y_continuous(limits = c(0, 70), breaks = seq(0, 70, by = 10)) +
  labs(x = "Middle Speed, mph",
       y = "Number of Vehicle")
p1
# ggsave(
#   filename = "images/chap11/ch11_spdHist.png",   plot = p1,   width = 6,   height = 4,   units = "in",   dpi = 300
# )

```


```{r}
#| label: freq dist graph


p2<- spdDataStat |> 
ggplot() +  theme_ipsum() +
  geom_rect(aes(xmin = spdLmtLower + 0.15, xmax = spdLmtUpper, ymin = 0, ymax = freq), alpha = 0.2)+
  geom_point(aes(x=spdMiddle, y= freq), color = "red") +
  geom_line(aes(x=spdMiddle, y= freq), group =1, color = "blue") +
  
  scale_x_continuous(limits = c(25, 70), breaks = seq(30, 70, by = 2)) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, by = 2)) +
  labs(x = "Middle Speed, mph",
       y = "Frequency (%)")

p2

# ggsave(
#   filename = "images/chap11/ch11_spdFreqPlot.png",   plot = p2,   width = 6,   height = 4,   units = "in",   dpi = 300
# )
```

```{r}
#| label: pdf graph


p2<- spdDataStat |>
  mutate(prob = freq / 2) |> 
ggplot() +  theme_ipsum() +
  geom_rect(aes(xmin = spdLmtLower + 0.15, xmax = spdLmtUpper, ymin = 0, ymax = freq), alpha = 0.2)+
  geom_point(aes(x=spdMiddle, y= freq), color = "red") +
  geom_line(aes(x=spdMiddle, y= freq), group =1, color = "blue") +
  geom_text(aes(x=spdMiddle, y= freq, label = freq), vjust = -0.5, hjust = -0.10, size = 4, angle = 90) +
  scale_x_continuous(limits = c(25, 70), breaks = seq(31, 70, by = 2)) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, by = 1)) +
  labs(x = "Middle Speed, mph",
       y = "Relative Frequency (%)")

p2

# ggsave(
#   filename = "images/chap11/ch11_spdFreqPlot.png",   plot = p2,   width = 6,   height = 4,   units = "in",   dpi = 300
# )
```


```{r}
#| label: cumul freq dist graph

spdDataStat |> 
ggplot() + theme_linedraw() +
  geom_point(aes(x=spdMiddle, y= cumulFreq)) +
  geom_line(aes(x=spdMiddle, y= cumulFreq), group =1) +
  scale_x_continuous(limits = c(25, 70), breaks = seq(25, 70, by = 5)) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 10)) +
  labs(x = "Middle Speed, mph",
       y = "Cumulative Frequency (%)")

```



# Summary Statistics: Dispersion

